---
title: "Indicator data: calculations and visualizations"
author: "Chloé R. Nater"
output: html_document
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NIcalc)
library(NIflex)
library(magrittr)
library(dplyr)
```

The basis for the Nature Index (NI) calculations are the indicator data. These are observations, model estimates, or expert judgements that quantify the state (typically abundance) of the different indicator species, species groups, or habitat types. 

The "raw" indicator data, including associated reference values, are stored in the NI database (https://naturindeks.nina.no/). As part of the presentation of NI2025, we want to present the following for each individual indicator: 

- Raw indicator values, by area, as time-series plots
- Raw indicator values, by area, as pair of value and uncertainty maps
- Scaled indicator values, by area, as time-series plots
- Scaled indicator values, by area, as pairs of value and uncertainty maps
- Area-averaged indicator values (referred to as "indicator index" in the documentation from NI2020), as part of the time-series plots for scaled indicator values

The area-averaged indicator values are also required to be fed back into the NI database as a separate table that can be used for corresponding visualizations on the webpage. 

## General setup

There are a few variables that are used repeatedly in the code in this chapter. 
We start off by defining these here: 

```{r}
## Indicator focal years
ind_focalYears <- c(1990, 1995, 2000, 2005, 2010, 2014, 2019, 2024)

## Toggle for using/ignoring reference value uncertainty
uncertain_refValue <- FALSE
#uncertain_refValue <- TRUE
```

Additionally, we will define and set a seed to ensure reproducibility of the simulations:
```{r}
mySeed <- 0
set.seed(mySeed)
```

## Connect to the Nature Index Database

To access and download indicator data, we need to retrieve a token from the NI database. 
For that to work, we need access to our NI database access credentials; we retrieve those from environmental variables (if you have not saved them as environmental variables, you can find the code for doing that in chapter 0; alternatively, you can also manually set them here):

```{r, eval = FALSE}
NIdb_username <- Sys.getenv("NIdb_username")
NIdb_password <- Sys.getenv("NIdb_password")
```

##  List indicators and retrieve data

If the person running this script has full access to the database (admin rights), they should be able to access the complete list of indicators by directly pulling it from the database using `NIcalc::getIndicators()`:

```{r, eval = FALSE}
indicatorList <- NIcalc::getIndicators() %>%
  dplyr::arrange(name)

saveRDS(indicatorList, file = "data/NI2025_IndicatorList.rds")
```

Alternatively, you can read in the saved version of the indicator list that comes with this repository: 
```{r}
indicatorList <- readRDS("data/NI2025_IndicatorList.rds")
```

The list is contains a total of 203 indicators, and is ordered alphabetically:
```{r}
indicatorList
```

Now that we have all indicators listed we can download the entire set of indicator data. This takes a while to run, and we therefore save the data import as a local file after it finished downloading so that we do not have to re-download the entire dataset repeatedly. 

```{r, eval = FALSE}
indicatorImport <- list()

for(i in 1:nrow(indicatorList)){
  indicatorImport[[i]] <- NIcalc::importDatasetApi(
    username = NIdb_username,
    password = NIdb_password,
    indic = indicatorList$name[i]
  )
}

names(indicatorImport) <- indicatorList$id

saveRDS(indicatorImport, "data/indicatorImport.rds")
```
```{r, echo = FALSE}
indicatorImport <- readRDS("data/IndicatorImport.rds")
```

The next step in the NI workflow is "assembling" the data. This step makes sure that the parameters characterizing uncertainty distributions for indicator values are present for all data points (see documentation of `NIcalc::assembleNiObject()` for more details). This also requires some time to run, and like the previous step, we will save the output so we can work directly from that afterwards.

```{r, eval = FALSE}
indicatorData <- list()

for(i in 1:nrow(indicatorList)){
  indicatorData[[i]] <- NIcalc::assembleNiObject(
    inputData = indicatorImport[[i]],
    predefNIunits = c(allArea = TRUE, parts = TRUE, counties = FALSE),
    indexType = "thematic"
  )
}

names(indicatorData) <- indicatorList$id

saveRDS(indicatorData, "data/indicatorData_assembled.rds")
```
```{r, echo = FALSE}
indicatorData <- readRDS("data/IndicatorData_assembled.rds")
```

## Simulate indicator value distributions

For visualizing indicator values (both original and scaled), we need associated measures of uncertainty. While all indicator value have such information associated in the database, the format thereof varies, i.e. some indicators have lower and upper quartiles in addition to means reported, while others are specified via custom distributions. 

We require a standardized way for presenting values including uncertainty across all indicators. 
In previous iterations of NI, it seems that results were displayed as median plus 95% confidence interval (= 2.5% and 97.5% quantiles, according to NINA Report 1990, "Nature Index System Documentation"). 
While working with alternative approaches to visualizing indicator data, we also experimented with using coefficient of variation (= relative standard deviation = sd/mean) for visualizing uncertainty in maps (see here https://ninanor.github.io/NIviz/maps-1.html), and displays of the entire indicator value distributions (e.g. 5.1 here: https://ninanor.github.io/NIviz/other-figures.html).

Based on this, we have a need for a) different measures of uncertainty and b) samples of draws from the uncertainty distributions of all indicators. We can obtain all of this information from simulations. The function `simulate_IndicatorData()` can be used to run the corresponding simulations for all indicators and years with available data: 

```{r}
## Source functions
source("R/simulate_IndicatorData.R")
source("R/scale_IndicatorValues.R")

## Make folder for storing data as files
if(!dir.exists("data/simulated_IndicatorData")){
  dir.create("data/simulated_IndicatorData")
}

## Set up list for storing all results
simData <- list()

## For all indicators, simulate data from distributions and save as files and in list

for(i in 1:nrow(indicatorList)){
  
  simData_i <- simulate_IndicatorData(indData = indicatorData[[i]])
  
  simData[[i]] <- simData_i
  
  saveRDS(simData_i, file = paste0("data/simulated_IndicatorData/simData_id_", indicatorList$id[i], ".rds"))
}

``` 


