---
title: 'Indicator data: calculations and visualizations'
author: "Chloé R. Nater"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(NIcalc)
library(NIflex)
library(magrittr)
library(dplyr)
library(ggplot2)
library(paletteer)
library(svglite)
```

The basis for the Nature Index (NI) calculations are the indicator data. These are observations, model estimates, or expert judgements that quantify the state (typically abundance) of the different indicator species, species groups, or habitat types. 

The "raw" indicator data, including associated reference values, are stored in the NI database (https://naturindeks.nina.no/). As part of the presentation of NI2025, we want to present the following for each individual indicator: 

- Raw indicator values, by area, as time-series plots
- Raw indicator values, by area, as pair of value and uncertainty maps
- Scaled indicator values, by area, as time-series plots
- Scaled indicator values, by area, as pairs of value and uncertainty maps
- Area-averaged indicator values (referred to as "indicator index" in the documentation from NI2020), as part of the time-series plots for scaled indicator values

The area-averaged indicator values are also required to be fed back into the NI database as a separate table that can be used for corresponding visualizations on the webpage. 

## General setup

There are a few variables that are used to control execution of parts of the code in this chapter. 
We start off by defining these here: 

```{r}
## Indicator focal years
ind_focalYears <- c(1990, 1995, 2000, 2005, 2010, 2014, 2019, 2024)

## Toggle for using/ignoring reference value uncertainty in visualizations
#uncertainRefValue <- TRUE
uncertainRefValue <- FALSE

## Toggle for visualizing scaled values truncated before or after summarising
truncAfterSummary <- TRUE
#truncAfterSummary <- FALSE
```

Additionally, we will define and set a seed to ensure reproducibility of the simulations:
```{r}
mySeed <- 0
set.seed(mySeed)
```

## Connect to the Nature Index Database

To access and download indicator data, we need to retrieve a token from the NI database. 
For that to work, we need access to our NI database access credentials; we retrieve those from environmental variables (if you have not saved them as environmental variables, you can find the code for doing that in chapter 0; alternatively, you can also manually set them here):

```{r, eval = FALSE}
NIdb_username <- Sys.getenv("NIdb_username")
NIdb_password <- Sys.getenv("NIdb_password")
```

After retrieving login credentials, we can open a connection / request a token: 
```{r, eval = FALSE}
NIcalc::getToken(username = NIdb_username,  
                 password = NIdb_password,
                 url = "https://www8.nina.no/NaturindeksNiCalc")
```
##  List indicators and retrieve data

If the person running this script has full access to the database (admin rights), they should be able to access the complete list of indicators by directly pulling it from the database using `NIcalc::getIndicators()`:

```{r, eval = FALSE}
indicatorList <- NIcalc::getIndicators() %>%
  dplyr::arrange(name)

saveRDS(indicatorList, file = "data/NI2025_IndicatorList.rds")
```

Alternatively, you can read in the saved version of the indicator list that comes with this repository: 
```{r}
indicatorList <- readRDS("data/NI2025_IndicatorList.rds")
```

The list is contains a total of 203 indicators, and is ordered alphabetically:
```{r}
indicatorList
```

Now that we have all indicators listed we can download the entire set of indicator data. This takes a while to run, and we therefore save the data import as a local file after it finished downloading so that we do not have to re-download the entire dataset repeatedly. 

```{r, eval = FALSE}
indicatorImport <- list()

for(i in 1:nrow(indicatorList)){
  indicatorImport[[i]] <- NIcalc::importDatasetApi(
    username = NIdb_username,
    password = NIdb_password,
    indic = indicatorList$name[i]
  )
}

names(indicatorImport) <- indicatorList$id

saveRDS(indicatorImport, "data/indicatorImport.rds")
```
```{r, echo = FALSE}
indicatorImport <- readRDS("data/IndicatorImport.rds")
```

The next step in the NI workflow is "assembling" the data. This step makes sure that the parameters characterizing uncertainty distributions for indicator values are present for all data points (see documentation of `NIcalc::assembleNiObject()` for more details). This also requires some time to run, and like the previous step, we will save the output so we can work directly from that afterwards.

```{r, eval = FALSE}
indicatorData <- list()

for(i in 1:nrow(indicatorList)){
  indicatorData[[i]] <- NIcalc::assembleNiObject(
    inputData = indicatorImport[[i]],
    predefNIunits = c(allArea = TRUE, parts = TRUE, counties = FALSE),
    indexType = "thematic"
  )
}

names(indicatorData) <- indicatorList$id

saveRDS(indicatorData, "data/indicatorData_assembled.rds")
```
```{r, echo = FALSE}
indicatorData <- readRDS("data/IndicatorData_assembled.rds")
```

## Simulate indicator value distributions

For visualizing indicator values (both original and scaled), we need associated measures of uncertainty. While all indicator value have such information associated in the database, the format thereof varies, i.e. some indicators have lower and upper quartiles in addition to means reported, while others are specified via custom distributions. 

We require a standardized way for presenting values including uncertainty across all indicators. 
In previous iterations of NI, it seems that results were displayed as median plus 95% confidence interval (= 2.5% and 97.5% quantiles, according to NINA Report 1990, "Nature Index System Documentation"). 
While working with alternative approaches to visualizing indicator data, we also experimented with using coefficient of variation (= relative standard deviation = sd/mean) for visualizing uncertainty in maps (see here https://ninanor.github.io/NIviz/maps-1.html), and displays of the entire indicator value distributions (e.g. 5.1 here: https://ninanor.github.io/NIviz/other-figures.html).

Based on this, we have a need for a) different measures of uncertainty and b) samples of draws from the uncertainty distributions of all indicators. We can obtain all of this information from simulations. The function `simulate_IndicatorData()` can be used to run the corresponding simulations for all indicators and years with available data: 

```{r, eval = FALSE}
## Source functions
source("R/simulate_IndicatorData.R")
source("R/scale_IndicatorValues.R")

## Make folder for storing data as files
if(!dir.exists("data/simulated_IndicatorData")){
  dir.create("data/simulated_IndicatorData")
}

## Set up list for storing all results
simData <- list()

## For all indicators, simulate data from distributions and save as files and in list

for(i in 1:nrow(indicatorList)){
  message("")
  message(crayon::bold(crayon::green(paste0(indicatorList$name[i]))))
  
  simData_i <- simulate_IndicatorData(indData = indicatorData[[i]])
  
  simData[[i]] <- simData_i
  
  saveRDS(simData_i, file = paste0("data/simulated_IndicatorData/simData_id_", indicatorList$id[i], ".rds"))
  message("")
}

## Save total list containing all datasets
saveRDS(simData, file = "data/simulated_IndicatorData/simData_all.rds")
``` 

## Summarize indicator value distributions

In the next step, we want to calculated some summary metrics for our indicator data. 
The helper function `summarise_IndicatorDist()` takes the raw and scaled indicator datasets and calculates the following metrics for each year- and area-specific indicator value: 

- Mean
- Median
- Standard deviation
- Relative standard deviation
- 2.5%, 5%, 25%, 75%, 95%, and 97.5% quantiles

We first source the function, and then run it on each of the three datasets (raw, scaled with fixed reference, scaled with sampled reference) available for all indicators. The results we will save as local .rds and .csv files. 

```{r, eval = FALSE}
## Source function
source("R/summarise_IndicatorDist.R")

## Make folder for storing data as files
if(!dir.exists("data/IndicatorData_StatSummaries")){
  dir.create("data/IndicatorData_StatSummaries")
}

savePath <- "data/IndicatorData_StatSummaries/"

## Re-set seed
set.seed(mySeed)

## Summarise indicator value distributions for all datasets and indicators
for(i in 1:nrow(indicatorList)){
  message(crayon::bold(crayon::green(paste0(indicatorList$name[i]))))
  
  # Raw (unscaled data)
  summarise_IndicatorDist(indSim = simData[[i]]$raw,
                          ind_id = indicatorList$id[i],
                          dataset_name = "raw",
                          savePath = savePath)
  
  # Scaled using fixed reference
  summarise_IndicatorDist(indSim = simData[[i]]$scaled_fixedRef,
                          ind_id = indicatorList$id[i],
                          dataset_name = "scaled_fixedRef",
                          savePath = savePath)
  
  # Scaled using sampled reference
  summarise_IndicatorDist(indSim = simData[[i]]$scaled_sampledRef,
                          ind_id = indicatorList$id[i],
                          dataset_name = "scaled_sampledRef",
                          savePath = savePath)
  message("")
}
```

## Calculating indicator indices

Traditionally, time-series of area-aggregated indicator values have been published for a selection of indicators on the NI webpage. These are referred to as "indicator indices" and are calculated using the same framework as the NI for different ecosystems (via `NIcalc::calculateIndex()`). We calculate these here using the wrapper function `calculate_IndicatorIndex()`. 
The source code underlying this function originates from scripts from NI2020 stored on NINA's internal servers under "R:/Prosjekter/12892000_naturindeks_rammeverk_database_og_innsynslosning/Naturindeks 2020/Beregninger/Indikatorer/". 
Back then, indicator indices were calculated on a per-ecosystem basis, likely to circumvent the issue that no alternative had been coded into "NIcalc" yet. In `calculate_IndicatorIndex()` here, we have made adjustments that allow calculating indicator indices taking into account all relevant ecosystems; in practice, this means that area weights are determined by the extent of area within each spatial unit covered by any of the ecosystems relevant to the indicator (as opposed to just one selected one). It's still possible to calculate indicator indices specific to single ecosystems too by setting `weigh_byEcosystem = TRUE` and specifying which ecosystem in the call to the function. However, for the purpose of visualizing indicator data, we will work with indicator indices calculated for all relevant ecosystems. 

Below, we loop over all indicators to calculate indicator indices, and save the results into a list for further use in R and as files into a designated folder: 

```{r, eval = FALSE}
## Source function
source("R/calculate_IndicatorIndex.R")

## Create folder for saving results
if(!dir.exists("results_Indicators")){
  dir.create("results_Indicators")
}

## Set up list for storing results
indIndex <- list()

## Calculate indicator indices and save results, per indicator
for(i in 1:nrow(indicatorList)){

  indIndex_i <- calculate_IndicatorIndex(dataSet = indicatorData[[i]],
                                     testRun = FALSE,
                                     NAimputation = FALSE,
                                     weigh_byEcosystem = FALSE,
                                     include_subAreas = TRUE,
                                     truncAtRef = TRUE,
                                     enforce_weightSum1 = TRUE)
  indIndex[[i]] <- indIndex_i
  
  readr::write_excel_csv(indIndex_i$summary, file = paste0("results_Indicators/indIndex_Summary_id_", indicatorList$id[i], ".csv"))
  
  saveRDS(indIndex_i, paste0("results_Indicators/indIndex_id_", indicatorList$id[i], ".rds"))
}


```

Note that we are calculating indicator indices without imputations (`NAimputation = FALSE`). This is because imputation using the "mice" algorithm will not work very well for many indicators due to limited data and high collinearity. It works for indices combining many indicators because that translates to more (variable) data. Without imputations, indicator indices for selected years represent area-weighted means of values for the areas with available data in the specified year. Since different areas may have data available for different years, changes in calculated indicator indices over time have to be interpreted with caution as changes may indicate a) changes in average indicator values and/or b) changes in composition of areas with values. We assist this visually by not "connecting" the indicator index estimates across years below. 

## Visualizing raw and scaled indicator data

For the remainder of this chapter, we will work with one indicator at a time and read in the respective simulated datasets anew. This means we can remove the large objects containing data for all indicators simultaneously from the workspace:
```{r, warning = FALSE}
rm(simData, indicatorData, indicatorImport, indIndex)
```

There are two types of visualizations of indicator data that we are aiming for:

1. Time-series plots of medians and 95% confidence intervals (featuring lines for different areas)
2. Year-specific maps of a) medians and b) uncertainty as relative standard deviation

Both visualizations will be done for the unscaled (raw) and scaled indicator values. 

Below, we present the corresponding visualizations in indicator-specific sub-chapters. There are a few things to note: 

- The defaults for visualization of scaled indicator values are using uncertainty statistics truncated after summarising, and treating the calculations based on fixed references as the baseline. This can be changed using the toggles `truncAfterSummary` and `uncertainRefValue` at the start of this script.
- Visualizations will be partially displayed in this document, and also saved as files.
- In this document, we will display 1. featuring all areas. We will not draw and display maps here as these will be part of the presentation of data and results on the new NI webpage.
- For now, the visualizations are static. The goal for the website is to have interactive elements in the figures, including: menus to toggle on and off different time series (and reference levels), "sliders" to change year for maps, simultaneous highlights in median and uncertainty maps, etc. 
- Visualizations of the entire value distributions (see under 5.1. here: https://ninanor.github.io/NIviz/other-figures.html) will not be done for individual indicators (only for the aggregated NI for the different ecosystems).

First, we set up directories, determine which scaled values file to use, define a plot color to use in cases where we have time series for many areas, and set up lists to store the plots in R:
```{r, eval = FALSE}
## Set up folder for plots (if not present already)
if(!dir.exists("plots")){
  dir.create("plots")
}

## Select correct file for scaled data
filename_scaled <- dplyr::case_when(
  !uncertainRefValue & truncAfterSummary ~ "fixedRef",
  !uncertainRefValue & !truncAfterSummary ~ "fixedRef_truncPreSum",
  uncertainRefValue & truncAfterSummary ~ "sampledRef",
  uncertainRefValue & !truncAfterSummary ~ "sampledRef_truncPreSum",
)

## Set fixed plotting colors

# Time-series plot color for indicators with > 20 areas
plot.col <- "#D37295FF"

# Time-series plot color for indicator indices
plot.col2 <- "#0B2120"

## Make lists to store plots for each indicator
TS_raw <- list()
TS_scaled <- list()
#MAP_raw <- list()
#MAP_scaled <- list()

```

Then, we proceed to plot time series and save plots to files: 
```{r, eval = FALSE}
for(i in 1:nrow(indicatorList)){
  
  ## Extract indicator name and id
  ind_id <- indicatorList$id[i]
  ind_name <- indicatorList$name[i]
  
  message(crayon::bold(crayon::green(paste0(ind_name))))
  
  ## Make subfolder for species in target directory & set save path
  if(!dir.exists(paste0("plots/id_", ind_id, "_", stringr::str_replace_all(ind_name, " ", "_")))){
    dir.create(paste0("plots/id_", ind_id, "_", stringr::str_replace_all(ind_name, " ", "_")))
  }
  savePath_i <- paste0("plots/id_", ind_id, "_", stringr::str_replace_all(ind_name, " ", "_"))
  
  ## Load indicator datasets
  message("Processing indicator data...")
  
  data_raw <- readRDS(paste0("data/IndicatorData_StatSummaries/statSummary_id_", ind_id, "_raw.rds")) %>%
    dplyr::mutate(year_t = as.integer(year))
  
  data_scaled <- readRDS(paste0("data/IndicatorData_StatSummaries/statSummary_id_", ind_id, "_scaled_", filename_scaled, ".rds")) %>%
    dplyr::mutate(year_t = as.integer(year))
  
  data_indIdx <- readRDS(paste0("results_Indicators/indIndex_id_", ind_id, ".rds"))$summary %>%
    dplyr::filter(indexArea == "wholeArea" & year_t %in% ind_focalYears & !data_singleArea ) %>%
    dplyr::mutate(year_t = as.integer(year_t))
  
  ## Truncate scaled values (if necessary)
  data_rows <- c("mean", "median", "q025", "q05", "q25", "q75", "q95", "q975") 
  
  data_scaled <- data_scaled %>%
    dplyr::mutate(dplyr::across(.cols = all_of(data_rows), ~ ifelse(. > 1, 1, .)))
  
  ## Count areas for picking time series display
  manyAreas <- ifelse(length(unique(data_raw$ICunitName)) > 20, TRUE, FALSE)
  
  ## Set years to highlight
  minYear <- min(data_raw$year_t)
  maxYear <- max(data_raw$year_t)
  
  if(minYear >= ind_focalYears[1]){
    plot.yearH <- ind_focalYears[which(ind_focalYears >= minYear)]
  }else{
    plot.yearH <- unique(c(minYear, rev(seq(from = ind_focalYears[1], to = minYear, by = -5)), ind_focalYears))
  }
  
  ## Prepare time series plots for each indicator
  message("Drawing time-series plots...")
  
  # Raw data
  if(!manyAreas){
    ts_raw <- ggplot(data_raw, aes(x = year_t)) + 
      geom_point(aes(y = median, color = ICunitName)) + 
      geom_errorbar(aes(ymin = q025, ymax = q975, color = ICunitName), width = 0.5, alpha = 0.25) + 
      geom_line(aes(y = median, color = ICunitName)) + 
      xlab("Year") + ylab("Indicator value (unscaled)") + 
      ggtitle(ind_name) +
      paletteer::scale_color_paletteer_d(palette = "ggthemes::Tableau_20", name = "Area") +
      scale_x_continuous(breaks = plot.yearH) + 
      theme_bw() + 
      theme(legend.position = "bottom", 
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(angle = 45, vjust = 0.8))
  }else{
    ts_raw <- ggplot(data_raw, aes(x = year_t)) + 
      geom_point(aes(y = median, group = ICunitName), color = plot.col, alpha = 0.2) + 
      geom_errorbar(aes(ymin = q025, ymax = q975, group = ICunitName), width = 0.5, color = plot.col, alpha = 0.1) + 
      geom_line(aes(y = median, group = ICunitName), color = plot.col, alpha = 0.2) + 
      xlab("Year") + ylab("Indicator value (unscaled)") + 
      ggtitle(ind_name) +
      scale_x_continuous(breaks = plot.yearH) + 
      theme_bw() + 
      theme(legend.position = "bottom", 
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(angle = 45, vjust = 0.8))
  }
  
  TS_raw[[i]] <- ts_raw
  ggsave(file = paste0(savePath_i, "/TimeSeries_raw.svg"), plot = ts_raw, width = 8, height = ifelse(manyAreas, 5, 6.5))
  ggsave(file = paste0(savePath_i, "/TimeSeries_raw.pdf"), plot = ts_raw, width = 8, height = ifelse(manyAreas, 5, 6.5))
  ggsave(file = paste0(savePath_i, "/TimeSeries_raw.png"), plot = ts_raw,width = 8, height = ifelse(manyAreas, 5, 6.5), dpi = 150)
  
  # Scaled data
  if(!manyAreas){
    ts_scaled <- ggplot(data_scaled, aes(x = year_t)) + 
      geom_point(aes(y = median, color = ICunitName)) + 
      geom_errorbar(aes(ymin = q025, ymax = q975, color = ICunitName), width = 0.5, alpha = 0.25) + 
      geom_line(aes(y = median, color = ICunitName)) + 
      geom_errorbar(data = data_indIdx, aes(x = year_t, ymin = q025, ymax = q975), width = 0.5, color = plot.col2) + 
      geom_point(data = data_indIdx, aes(x = year_t, y = median), shape = 21, size = 2, stroke = 1.5, fill = "white", color = plot.col2) +
      ylim(0, 1.01) + 
      xlab("Year") + ylab("Indicator value (scaled)") + 
      ggtitle(ind_name) +
      paletteer::scale_color_paletteer_d(palette = "ggthemes::Tableau_20", name = "Area") + 
      scale_x_continuous(breaks = plot.yearH) + 
      theme_bw() + 
      theme(legend.position = "bottom", 
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(angle = 45, vjust = 0.8))
  }else{
    ts_scaled <- ggplot(data_scaled, aes(x = year_t)) + 
      geom_point(aes(y = median, group = ICunitName), color = plot.col, alpha = 0.2) + 
      geom_errorbar(aes(ymin = q025, ymax = q975, group = ICunitName), width = 0.5, color = plot.col, alpha = 0.1) + 
      geom_line(aes(y = median, group = ICunitName), color = plot.col, alpha = 0.2) +
      geom_errorbar(data = data_indIdx, aes(x = year_t, ymin = q025, ymax = q975), width = 0.5, color = plot.col2) + 
      geom_point(data = data_indIdx, aes(x = year_t, y = median), shape = 21, size = 2, stroke = 1.5, fill = "white", color = plot.col2) +
      ylim(0, 1) + 
      xlab("Year") + ylab("Indicator value (scaled)") + 
      ggtitle(ind_name) +
      scale_x_continuous(breaks = plot.yearH) + 
      theme_bw() + 
      theme(legend.position = "bottom", 
            panel.grid.minor = element_blank(),
            axis.text.x = element_text(angle = 45, vjust = 0.8))
  }
  
  TS_scaled[[i]] <- ts_scaled
  ggsave(file = paste0(savePath_i, "/TimeSeries_scaled.svg"), plot = ts_scaled, width = 8, height = ifelse(manyAreas, 5, 6.5))
  ggsave(file = paste0(savePath_i, "/TimeSeries_scaled.pdf"), plot = ts_scaled, width = 8, height = ifelse(manyAreas, 5, 6.5))
  ggsave(file = paste0(savePath_i, "/TimeSeries_scaled.png"), plot = ts_scaled, width = 8, height = ifelse(manyAreas, 5, 6.5), dpi = 150)
  
  message("")
}

## Save entire plot lists for re-use later
saveRDS(TS_raw, file = "results_Indicators/PlotList_TimeSeries_raw.rds")
saveRDS(TS_scaled, file = "results_Indicators/PlotList_TimeSeries_scaled.rds")
```
```{r, echo = FALSE, eval = FALSE}
TS_raw <- readRDS("results_Indicators/PlotList_TimeSeries_raw.rds")
TS_scaled <- readRDS("results_Indicators/PlotList_TimeSeries_scaled.rds")
```

## Indicator presentation

Below, we show the time series plots for raw and scaled indicator data. 

```{r, results = "asis", echo = FALSE, eval = TRUE}
## Plot display for HTML render
for(i in 1:nrow(indicatorList)){

  ind_id <- indicatorList$id[i]
  ind_name <- indicatorList$name[i]
  
  savePath_i <- paste0("plots/id_", ind_id, "_", stringr::str_replace_all(ind_name, " ", "_"))

  #raw_file <- paste0(savePath_i, "/TimeSeries_raw.svg")
  #scaled_file <- paste0(savePath_i, "/TimeSeries_scaled.svg")
  
  raw_file <- paste0(savePath_i, "/TimeSeries_raw.png")
  scaled_file <- paste0(savePath_i, "/TimeSeries_scaled.png")
  
  cat("\n\n### ", indicatorList$name[i], "\n\n")
  cat("![](", raw_file, ")\n\n")
  cat("![](", scaled_file, ")\n\n")
}

```

```{r, results = "asis", echo = FALSE, eval = FALSE}
## Plot display for PDF render
for(i in 1:nrow(indicatorList)){

  ind_id <- indicatorList$id[i]
  ind_name <- indicatorList$name[i]
  
  savePath_i <- paste0("plots/id_", ind_id, "_", stringr::str_replace_all(ind_name, " ", "_"))

  raw_file <- paste0(savePath_i, "/TimeSeries_raw.pdf")
  scaled_file <- paste0(savePath_i, "/TimeSeries_scaled.pdf")
  
  cat("\n\n### ", indicatorList$name[i], "\n\n")
  cat("![](", raw_file, "){height=4in}\n\n")
  cat("![](", scaled_file, "){height=4in}\n\n")
}

```